TacticToe: Learning to Prove with Tactics

### 2018-02-19 tactictoe_jv.tex
minor
Sec 11

\paragraph{Minimizing length of the proof}
A fast and simple minimization is applied when processing the final proof. If
the tactic \texttt{A THEN B} appears in the proof and has the same effect as
\texttt{B} then we can replace \texttt{A THEN B} by \texttt{B} in the script.
Optionally, stronger minimization can be obtained
by rerunning \tactictoe with a low prior
policy coefficient, no prior evaluation and a database of tactics that contains
tactics from the discovered proof only.

\paragraph{Minimizing tactic arguments}
Let $t$ be a tactic applied to a goal $g$ containing a list $l$ (of theorems)
as argument. And let $t'$ be the tactic $t$ where one element $e$ of $l$ as be
removed. If $t$ and $t'$ have the same effect on $g$ then $t'$ can replace $t$
in the final proof. This process is repeated for each element of $l$.
This is a generalization of the simplest method used for minimizing a list of
theorems in ``hammers'' \cite{hammers4qed}.

\paragraph{Prettification}
Without prettification, the returned tactic is barely readable as it contains
information to guarantees that each \sml subterm is interpreted in the way in
any context. Since we return the proof at a point where the \sml interpreter is
a specific state, we can strip unnecessary information in the form of
module prefixes. If possible, we group all local declarations in the
script under a single \texttt{let} binding at the start of the script. And we
replace extended terms by their quoted version.
All in all, if a prettified tactic $t_p$ has the same effect its original
$t_o$, we replace $t_p$ by $t_o$ in the proof.\\

@@

\section{Case Study}\todo{comment on the case study}
Since the proofs generated by \tactictoe are meant to be part of a \holfour development, it interesting to compare their quality with the original human proofs.
The quality of a proof in \holfour can be measured in terms of length, readability, maintainability and verification speed.
We study these properties in three examples taken from our full-scale experiment.
The time needed by \holfour to check a proof is shown in parentheses.

Example~\ref{ex:cs1} is about proving that greatest common divisors are unique.
The human developer recognizes that it follows easily from two theorems.
Our prover \tactictoe is not able to find that \texttt{DIVIDES\_ANTISYM} is
necessary directly from the shape of the original theorems so it rewrites the
goal then splits the goal into multiple cases until the goal obligations are similar
to \texttt{DIVIDES\_ANTISYM} to be predicted. As expected, \tactictoe's proof
also takes much longer to check.

@@

In Example~\ref{ex:cs3}, we prove a theorem about lists, a domain where \tactictoe excels compared to ATPs.
Given two list $l_1$ and $l_2$ where the length of $l_1$ (noted $p$) is less
than a natural
number $n$, the theorem states updating the $n^{th}$
element of the concatenation of $l_1$ and $l_2$ is the same as
updating the $m^{th}$ element of $l_2$ where $m = n - p$. Again,
\tactictoe's proof is
much more readable starting with applying induction on $l$. It solves the base
case of the induction by rewriting and distinguish cases where $n=0$ or $n>0$
in the inductive hypothesis. It finalizes the proof with a short call to \metis
using the definition of
\texttt{LUPDATE}. Here, \tactictoe's proof is arguably smaller, faster and
easier to
understand and maintain. Such proofs after an expert review could replace the
their respective original human proofs in the \holfour repository.

@@

In Example~\ref{ex:cs4}, we demonstrate the capabilities of \tactictoe on a
goal that was not already proved in the \holfour libraries: that the set of
numbers $\{0,...,n+m-1\} \setminus \{0,...,n-1\}$ is the same as the set
obtained by adding $n$ to everything in $\{0,...,m-1\}$. In this example,
\tactictoe uses the simplification set \texttt{ARITH\_ss} to reduce arithmetic
formulas. This exemplifies an another advantage that \tactictoe has over ATPs,
mainly its ability to take advantage of user-defined simplification sets.

>>>>

\paragraph{Proof Length Minimization}
A fast and simple minimization is applied when processing the final proof. If
the tactic \texttt{A THEN B} appears in the proof and has the same effect as
\texttt{B} then we can replace \texttt{A THEN B} by \texttt{B} in the script.
Optionally, stronger minimization can be obtained
by rerunning \tactictoe with a low prior
policy coefficient, no prior evaluation and a database of tactics that contains
tactics from the discovered proof only.

\paragraph{Tactic Arguments Minimization}
Let $t$ be a tactic applied to a goal $g$ containing a list $l$ (of theorems)
as argument. And let $t'$ be the tactic $t$ where one element $e$ of $l$ as be
removed. If $t$ and $t'$ have the same effect on $g$ then $t'$ can replace $t$
in the final proof. This process is repeated for each element of $l$.
This is a generalization of the simplest method used for minimizing a list of
theorems in ``hammers'' \cite{hammers4qed}.

\paragraph{Prettification}
Without prettification, the returned tactic is barely readable as it contains
information to guarantees that each \sml subterm is interpreted in the way in
any context. Since we return the proof at a point where the \sml interpreter is
in a specific state, we can strip unnecessary information, such as
module prefixes. If possible, we group all local declarations in the
script under a single \texttt{let} binding at the start of the script. We
also replace extended terms by their quoted version.
All in all, if a prettified tactic $t_p$ has the same effect its original
$t_o$, we replace $t_p$ by $t_o$ in the proof.\\

@@

\section{Case Study}\todo{comment on the case study}
Since the proofs generated by \tactictoe are meant to be parts of a \holfour development, it interesting to compare their quality with the original human proofs.
The quality of a proof in \holfour can be measured in terms of length, readability, maintainability and verification speed.
We study these properties in three examples taken from our full-scale experiment.
We list the time needed by \holfour to check a proof in parentheses.

We start with Example~\ref{ex:cs1}, which proves that greatest common divisors are unique.
The human formalizer recognized that it follows from two theorems.
The relevance filtering of \tactictoe is not as powerful as that used in hammers
and is therefore not able to find the \texttt{DIVIDES\_ANTISYM} property. The proof
proceeds instead by rewriting the
goal with the definitions then spliting the goal into multiple cases until the goal obligations are similar
to \texttt{DIVIDES\_ANTISYM}. As expected, \tactictoe's proof
also takes much longer to check.

@@

In Example~\ref{ex:cs3}, we prove a theorem about lists, a domain where \tactictoe excels compared to ATPs.
Given two list $l_1$ and $l_2$ where the length of $l_1$ (noted $p$) is less
than a natural
number $n$, the theorem states updating the $n^{th}$
element of the concatenation of $l_1$ and $l_2$ is the same as
updating the $m^{th}$ element of $l_2$ where $m = n - p$. Again, the
\tactictoe's proof is
much more readable. It starts by applying induction on $l$. It solves the base
case of the induction by rewriting and proceeds by cases on $n=0$ or $n>0$
in the induction hypothesis. It finalizes the proof with a short call to \metis
using the definition of
\texttt{LUPDATE}. Here, \tactictoe's proof is arguably smaller, faster and
easier to
understand and maintain. Such proofs after an expert review could replace the
their respective original human proofs in the \holfour repository.

@@

In Example~\ref{ex:cs4}, we experiment with \tactictoe on a
goal that does not originate from the \holfour library. The conjecture is
that the set of
numbers $\{0,...,n+m-1\} \setminus \{0,...,n-1\}$ is the same as the set
obtained by adding $n$ to everything in $\{0,...,m-1\}$. In this example,
\tactictoe uses the simplification set \texttt{ARITH\_ss} to reduce arithmetic
formulas. This exemplifies another advantage that \tactictoe has over ATPs,
namely its ability to take advantage of user-defined simplification sets.
